\section*{Letter of Transmittal}
\addcontentsline{toc}{section}{Letter of Transmittal}
\thispagestyle{plain}

\begin{samepage}
\begin{singlespace}

\noindent
\begin{minipage}[t]{0.48\textwidth}
{\bfseries California Polytechnic State University}\\
Biological Sciences Department\\
1 Grand Avenue\\
San Luis Obispo, CA 93407
\end{minipage}%
\hfill
\begin{minipage}[t]{0.48\textwidth}
\raggedleft
Emma Pelton\\
The Xerces Society for Invertebrate Conservation\\
628 NE Broadway, Suite 200\\
Portland, OR 97232
\end{minipage}

\vspace{1.5em}

\noindent December 31, 2025

\vspace{1.5em}

\noindent\textbf{Re: Final Technical Report -- Subaward CalPol23-4132}

\vspace{0.75em}

\noindent Dear Ms.\ Pelton,

\vspace{0.75em}

\noindent This report constitutes the final technical deliverable under Subaward CalPol23-4132, ``Wind Model Validation for Monarch Butterfly Overwintering Groves,'' funded through USDA Forest Service Award 21-DG-11132762-274 via The Xerces Society for Invertebrate Conservation.

The original statement of work proposed computational fluid dynamics (CFD) simulations using LiDAR-derived forest geometry to model wind patterns in overwintering groves. During the first year of the project, the drone-mounted LiDAR platform required for aerial data acquisition was damaged during training operations and remained out of service for an extended period, making the originally proposed methodology infeasible within the project timeline. In coordination with the Xerces Society, the research approach was revised to pursue direct empirical investigation of wind effects on overwintering monarch butterfly clusters at Vandenberg Space Force Base.

The revised study deployed time-lapse cameras and wind sensors at monarch cluster roost sites over the 2023--2024 overwintering season, producing 1,894 paired observations across 78 days. This approach enabled the first direct empirical test of the Disruptive Wind Hypothesis, a foundational assumption in monarch habitat management for over three decades. The findings have direct implications for habitat management practices at overwintering sites managed by the Department of Defense, U.S.\ Forest Service, California State Parks, and other land managers.

The enclosed report presents the complete methodology, results, and analysis in publication-quality format as specified in the statement of work. Questions regarding this report may be directed to Francis Villablanca, Ph.D., Professor Emeritus of Biology, California Polytechnic State University, San Luis Obispo, CA 93407.

\vspace{1.5em}

\noindent Sincerely,

\vspace{2em}

\noindent Francis Villablanca, Ph.D.\\
Principal Investigator\\
Professor Emeritus of Biology\\
California Polytechnic State University

\end{singlespace}
\end{samepage}
