\section{Materials and Methods}

\subsection{Study Site}

Site selection followed a systematic filtering process driven by project requirements and practical constraints. The study was supported by a federal grant that mandated research be conducted on federal lands. We selected Vandenberg Space Force Base (VSFB, 34.7398°N, 120.5725°W) in Santa Barbara County, California, based on several key advantages: mild winters with infrequent frost events, extensive historical plantings of blue gum eucalyptus (\textit{Eucalyptus globulus}) that have created suitable overwintering habitat throughout the installation, and restricted access that provided security for long-term equipment deployment. The base contains thirty documented monarch overwintering groves, with several sites consistently ranking within the top 10\% of population counts statewide over the past decade \autocite{xercesGuideWesternMonarch2025}. 

Working with the base's monarch conservation coordinator, we initially screened twelve locations from the thirty sites based on their documented capacity to support monarch aggregations and provide year-round access. This collaboration leveraged local expertise from managing Western Monarch Thanksgiving Count activities for multiple years \autocite{xercesGuideWesternMonarch2025}. During the study period, ten of these sites were actively monitored. However, due to low monarch populations during the 2023-2024 season and no observed overwintering behavior in the 2024-2025 season, only two sites—Spring Canyon and UDMH—produced measurable butterfly clusters suitable for our analysis.

Spring Canyon (34.6315°N, 120.6182°W) represents the most productive and historically reliable overwintering site on VSFB. Located in South Base within 300 meters of Space Launch Complex 4, this approximately 2.0-hectare site consists entirely of mature blue gum eucalyptus trees reaching heights of approximately 40 meters. An unnamed perennial creek runs through the center of the grove, creating a riparian corridor that supports heterogeneous canopy structure with variable tree spacing and diverse understory vegetation. Surf Road, an infrequently used paved access road, bisects both the perennial creek and forest canopy. 

The UDMH site (34.6719°N, 120.5950°W), also located in South Base, comprises a 5.1-hectare eucalyptus grove planted in windrows adjacent to a waste treatment facility. The uniformly spaced trees maintain a largely clear understory with scattered low shrubs. Although only recently documented as an overwintering location in 2022, UDMH immediately emerged as a significant site, supporting over 6,000 monarchs during its initial count and ranking among the base's highest population sites.

% [Figure here showing all thirteen groves, with different symbology for the various subsets]

\subsection{Monitoring Strategy}

Equipment deployment strategies differed between monitoring seasons to accommodate research objectives and field experience. During the 2023-2024 season, we employed two strategies: targeted deployments at sites with confirmed monarch presence, and anticipatory deployments at locations where monarchs were expected based on historical data but not currently observed. Targeted deployments concentrated at Spring Canyon and UDMH where active aggregations were documented throughout the season. Anticipatory deployments occurred at four overwintering sites—additional locations within Spring Canyon and UDMH, plus SLC-6 and Tangair. No monarchs were recorded at anticipatory deployment sites; consequently, these data are excluded from analysis.

For the 2024-2025 season, we modified our approach to establish monitoring stations at ten sites before monarch arrival, based on historical occurrence records compiled by the base conservation coordinator. This expanded spatial coverage aimed to capture greater environmental variation across potential overwintering sites. However, the 2024-2025 season coincided with historically low monarch abundance throughout California \autocite{xerces_society_western_2025}, resulting in no observed clustering behavior at any monitored location on base. Consequently, our final dataset comprises two sites—Spring Canyon and UDMH—from the 2023-2024 season only.

\subsection{Field Equipment}

To observe changes in monarch abundance in response to strong wind events, we deployed remote monitoring equipment near butterfly clusters at overwintering sites. Field observations utilized 15-meter telescoping fiberglass poles (Max-Gain Systems, Inc., Marietta, GA) anchored at three points using ground anchors with guy lines securing both the top and base to create stable, freestanding structures. 

Poles were positioned 4-17 meters from cluster locations. This range, determined through field testing, balanced image resolution requirements for our grid-based counting method against disturbance minimization. Closer positioning compromised field of view, while greater distances degraded butterfly visibility below classification thresholds. Pole placement considered ground stability for the 15-meter structures, infrastructure clearance requirements, and clear viewing angles. When deploying near active clusters, we approached from directions that minimized disturbance; no butterfly dispersal was observed during equipment deployment. 

We monitored monarch abundance using modified trail cameras (GardePro E7 and E8, Shenzhen, China) configured for near-infrared imaging to enhance contrast between clustering butterflies and surrounding vegetation. Trail cameras were selected for their durability in extended field deployment, native time-lapse functionality, and modification potential. Near-infrared wavelength selection followed previous literature demonstrating effectiveness for butterfly population estimation \autocite{hristov_estimating_2019}. 

Hardware modifications exploited the camera's internal filter-switching mechanism by engaging nighttime mode to access the clear glass filter position, then disconnecting power to prevent reversion to the infrared cut filter. Near-infrared pass filters (>850 nm) were mounted externally to restrict incoming light to NIR wavelengths. This configuration produced images where clustering butterflies appeared as dark masses against bright eucalyptus foliage reflectance in the near-infrared spectrum. Field validation confirmed sufficient contrast for visual distinction of monarch clusters from background vegetation, supporting our human-labeler analytical approach.

Cameras were mounted atop poles using lightweight tie-down straps and positioned horizontally toward butterfly clusters at roosting height. The wireless live view feature enabled real-time preview and precise camera aiming during deployment. Cameras operated in time-lapse mode with motion detection disabled. 

Sampling interval selection balanced temporal resolution, battery life, and data processing feasibility through empirical optimization and rigorous statistical validation. Initial deployments used 10-minute intervals to capture significant changes in butterfly abundance, which preliminary observations indicated occurred on hourly rather than minute scales, while maintaining approximately 6-week continuous operation. Post-deployment statistical analysis using mixed-effects models and information-theoretic approaches systematically compared multiple sampling intervals across deployments. We conducted sequential subsample analyses starting with full temporal resolution and progressively testing reduced frequencies. Information-theoretic model comparison using Akaike Information Criterion (AIC) demonstrated that 30-minute intervals provided optimal balance—losing less than 5\% of information compared to full temporal resolution (measured by root mean square error) while reducing image classification workload by 67\%. Variance comparison analysis and visual assessment of fitted trend lines confirmed that this interval preserved essential time-series patterns including diurnal activity cycles, weather-response dynamics, and multi-day population trends. Battery life constraints and field deployment logistics further supported this interval choice, enabling extended autonomous operation essential for capturing complete behavioral sequences during variable weather conditions.

Wind monitoring equipment consisted of Rain Wise WindLog Wind Data Loggers (Rain Wise Inc., Trenton, Maine) installed at pole apices to measure wind at heights approximating butterfly roosting locations. These instruments recorded average wind speed and maximum wind gust at one-minute intervals—the highest frequency supported by the sensors. This recording interval enabled calculation of wind speed variance within each photographic sampling period, capturing gustiness lost with longer averaging periods. 

To systematically organize our heterogeneous monitoring efforts, we defined discrete monitoring periods as deployment units. Each deployment represented a unique combination of monitoring location, camera configuration (including camera ID, mounting height, and viewing angle), associated wind measurements, and temporal coverage period. Since equipment was frequently reused across locations and time periods, this deployment-based structure provided standardized sampling units that accounted for variation in environmental conditions and equipment configurations while treating each deployment as independent for statistical analyses. This approach produced time-series images from each deployment for estimating monarch cluster abundance through systematic grid-based counting methods, enabling analysis of abundance patterns in relation to wind speed and other environmental variables.

% [A tree with a measuring device and a diagram AI-generated content may be incorrect.]

\subsection{Image Analysis}

\subsubsection{Grid-based Counting Method}

To quantify changes in monarch butterfly abundance from collected imagery, we developed a systematic grid-based counting protocol balancing accuracy with the practical constraints of analyzing tens of thousands of images. This approach addressed the challenge of estimating abundance in large aggregations where individual counts would be prohibitively time-consuming and emulated field researcher methods, including those used in the annual Thanksgiving Count \autocite{xercesProtocolsWesternMonarch2017}. We subdivided each image using a grid overlay system where human labelers assigned order-of-magnitude estimates per cell. Grid dimensions remained fixed throughout each deployment to ensure consistency. Custom software developed using the Electron framework in JavaScript facilitated this labeling effort.

Grid cell size varied by deployment based on camera-to-cluster distance. Cell dimensions were optimized to ensure most occupied cells contained butterflies in the 10–99 count range, balancing classification efficiency with spatial resolution. This standardization minimized cells alternating between widely different order-of-magnitude categories across the time series.

\subsubsection{Counting Protocol}

Human labelers estimated butterfly abundance within each grid cell using four order-of-magnitude categories: 0 (no butterflies), 1–9 (single digits), 10–99 (dozens), and 100–999 (hundreds). Labelers trained using a comprehensive online guide with example images and detailed classification criteria (\url{https://kylenessen.github.io/monarch_trailcam_classifier/}). The protocol prioritized efficiency while maintaining consistency across observers.

Because abundance estimates derived exclusively from two-dimensional photographic images, our classification protocol quantified only butterflies visible in the image plane without estimating three-dimensional cluster structure or depth. This approach intentionally excluded hidden individuals behind visible butterflies in overlapping aggregations, providing a conservative but consistent measure reflecting observable surface area rather than total volume. For cells containing partial butterflies at grid boundaries, labelers included these in counts unless double-counting would cause an adjacent cell to move to a higher category. When butterfly counts fluctuated between categories across the time series, we consistently applied the lower estimate to maintain conservative abundance estimates.

In addition to estimating monarch abundance, labelers recorded whether cells received direct sunlight. Direct sunlight classification presented challenges because oversaturated conditions eliminated the contrast enabling butterfly detection in shaded areas. Labelers classified cells as receiving direct sunlight when branches or butterflies exhibited additional illumination clearly from direct rather than indirect light, even when individual butterflies became difficult to distinguish due to pixel oversaturation. This classification required careful attention to subtle shape recognition and contextual awareness about butterfly locations established from previous images in the time series. This measurement was recorded only for occupied cells and stored separately.

Labelers received ongoing feedback throughout the classification process. All classifications underwent review for common errors including mislabeled cells, incorrect category assignments, and inconsistent counting criteria application. Direct communication of corrections to labelers ensured consistent protocol application.

\subsubsection{Abundance Calculation}

We calculated an abundance index for each frame by summing the products of cell counts and their assigned category values across all grid cells, employing conservative estimates using minimum values within each order-of-magnitude category:

\begin{equation}
\text{Abundance index} = \sum_{i} \rho_i \times C_i
\end{equation}

where $\rho_i$ represents the number of cells in category $i$, and $C_i$ represents the conservative estimate for that category. We used minimum category values ($C_1 = 1$ for category 1–9, $C_2 = 10$ for category 10–99, and $C_3 = 100$ for category 100–999) rather than midpoint or maximum values to ensure temporal analyses reflected genuine population shifts rather than estimation uncertainty.

\subsection{Statistical Analysis}
\label{sec:statistical-analysis}

To test our hypotheses, we developed a sequential, three-step analytical framework. This approach was designed to first assess the specific, established 2 m/s wind speed threshold for disruptive wind (Leong 2016) and then, if the data did not support this specific threshold, to explore the relationship between wind and monarch abundance more broadly.

\subsubsection{Step 1: Direct Test of the 2 m/s Threshold}

Our first step is a direct test of Leong's hypothesis using Generalized Linear Mixed-Effects Models (GLMMs). To properly model our count-based abundance index and account for likely overdispersion, we will use a negative binomial distribution. The response variable will be the `abundance_index` at time t, a metric representing the sum of conservative butterfly estimates within the sampled image. To account for temporal dependency, we will include the `abundance_index` from the previous time step (t-1) as a covariate. We will create two predictor variables representing the number of minutes that sustained and gust wind speeds exceed the 2 m/s threshold and test them in separate models to avoid issues with collinearity.

\begin{verbatim}
% Load required library
library(lme4)

# Model 1: Test sustained wind threshold with Negative Binomial GLMM
model_sustained_threshold <- glmer(
    abundance_index_t ~ abundance_index_t_minus_1 + 
                        sustained_minutes_above_2ms + 
                        sunlight_exposure_prop * ambient_temp +
                        (1 | view/labeler),
    family = "nbinom2",
    data = monarch_data
)

# Model 2: Test gust wind threshold with Negative Binomial GLMM
model_gust_threshold <- glmer(
    abundance_index_t ~ abundance_index_t_minus_1 + 
                        gust_minutes_above_2ms + 
                        sunlight_exposure_prop * ambient_temp +
                        (1 | view/labeler),
    family = "nbinom2",
    data = monarch_data
)
\end{verbatim}

If either analysis reveals a significant negative relationship, it would provide strong empirical support for the 2 m/s threshold. If not, as we suspect, it would suggest the relationship is more complex, leading us to the next step.

\subsubsection{Step 2: General Wind-Abundance Relationship Model}

If the specific 2 m/s threshold is not a strong predictor, we will broaden our inquiry to test whether wind, in a more general sense, influences monarch abundance. While I initially wanted to model the change in abundance directly, this approach creates a response variable with a complex and non-standard statistical distribution, making model interpretation unreliable. To overcome this, I propose we use a more robust method by modeling the abundance_index at each time step directly within a Generalized Linear Mixed-Effects Model (GLMM) framework, using a negative binomial distribution that is commonly used for count data.

A key challenge in analyzing time-series data is that the number of butterflies at any given moment is highly dependent on the number present just before. To properly account for this, our model will include the abundance from the previous time step (abundance_index_t_minus_1) as a predictor variable. By doing so, the model effectively asks: "Given the starting number of butterflies, what is the additional influence of our environmental predictors (e.g., wind, sun, temperature) on the abundance in the current time step?" This method allows us to statistically isolate the factors driving change between observations without creating a problematic response variable. It serves as more robust alternative for accounting autocorrelation than AR1.

To identify the most relevant aspect of wind, we will test a suite of continuous wind metrics, mean wind speed (sustained wind), maximum wind speed (peak gusts), and wind speed variance (gustiness), in separate models to avoid issues with collinearity. We will then use a model selection criterion, such as the Akaike Information Criterion (AICc), to compare these models. To ensure valid comparisons, the random effects structure will be held constant across all models. This entire analytical process will be repeated across multiple temporal scales (e.g., 30-minute, 4-hour, and daily) to investigate whether the disruptive effects of wind are due to immediate behavioral responses to acute events or cumulative responses to sustained conditions over longer durations. These scales were chosen to capture a range of potential biological responses, from short-term behavioral adjustments (30-minute) to responses to prolonged weather patterns (4-hour and daily), though the final scales may be adjusted based on data exploration.

\begin{verbatim}
# Example model for one wind metric (run separately for each)
model_max_wind <- glmer(
    abundance_index_t ~ abundance_index_t_minus_1 + 
                        max_wind_speed + 
                        sunlight_exposure_prop * ambient_temp +
                        (1 | view/labeler),
    family = "nbinom2",
    data = monarch_data
)
\end{verbatim}

If this general model reveals that one or more wind metrics are significant predictors, we will proceed to the final step.

\subsubsection{Step 3: Predictive Roost Abandonment Model}

If we establish a significant relationship in Step 2, our final step will be to build a predictive roost abandonment model. Using a mixed-effects logistic regression model, we will predict the probability of a roost abandonment event (e.g., a drop in abundance exceeding a high threshold, such as >90\%, to be finalized after data exploration), coded as a binary variable `roost_abandoned`. This model will use the most influential wind metric from Step 2 as the primary predictor.

\begin{verbatim}
% Load required library
library(lme4)

# Model: Predictive Hazard Analysis
hazard_model <- glmer(
    roost_abandoned ~ max_wind_speed + 
                      sunlight_exposure_prop * ambient_temp,
    family = binomial, # Specifies logistic regression
    data = monarch_data,
    control = glmerControl(optimizer = "bobyqa")
)
\end{verbatim}

The goal is to produce a probabilistic curve that quantifies the risk of roost abandonment across a range of wind speeds, providing a nuanced tool for habitat managers.
