% if only have one acknowledgment then uncomment this line
%\renewcommand{\acknowledgename}{Acknowledgment}

\begin{acknowledgments}

    I would like to express my sincere gratitude to my committee for their mentorship, support, and advocacy throughout this thesis. In particular, I thank Dr. Francis Villablanca, who gave his time generously by providing invaluable mentorship on the scientific process.

    This research benefited immensely from the ``Super Study'' group, whose many meetings provided excellent feedback and insights at all stages of my project. I thank Emma Pelton, Peter Ibsen, Jay Diffendorfer, Jessica Griffiths, Ashley Fisher, Sara Cuadra-Vargas, Charis van der Heide, Stephanie Little, and Ray Danner for their collaboration and expertise. Emma Pelton was instrumental in gathering these experts together, securing funding for my project, and initially seeding the idea for this study. Peter Ibsen and Jay Diffendorfer at USGS provided crucial funding support and valuable feedback throughout. Jessica Griffiths was instrumental in securing site access at Vandenberg Space Force Base, helping select study sites, and interpreting results. Ashley Fisher provided extensive discussions about monarch ecology and field assistance.

    I am deeply grateful to my undergraduate research assistants who conducted all photo labeling and provided essential field assistance: Skyler Meinholz, Vincent Rios, Emery Doornbos, and Kevin Untalan. Their dedication and hard work made this research possible.

    I thank LynneDee Althouse, Dan Meade, Stu Weiss, and Charis van der Heide, along with the many wonderful people at Althouse and Meade, Inc., for their collaboration and mentorship during my early monarch work, which laid the foundation for this thesis.

    At Cal Poly, I thank my graduate advisor Dr. Sean Lema, administrative staff Kristin Reeves and Katie Doctor, and my cohort members. I am grateful to Gage Willey, the Plant Conservatory Curator, who provided both office space and friendship throughout my master's.

    I thank my family and friends for their unwavering support and encouragement as I changed careers to pursue this project, especially my wife, Emily Nessen, whose patience and encouragement sustained me throughout.

    This research was made possible through generous funding from The Xerces Society for Invertebrate Conservation, US Forest Service International Programs, Department of Defense Legacy Resource Management Program, and the United States Geological Survey.

\end{acknowledgments}
